Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review

### 2017-12-15 analysis/paper/paper.Rmd
changes small
finished correction from Prof Worm

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/template.docx"
bibliography: references.bib
csl: mee.csl
abstract: |
  **Introduction:**  Refractory anaphylaxis cases do not respond to treatment with intramuscular adrenaline. The incidence of refractory anaphylaxis is estimated to range between 3--5% of all cases of anaphylaxis. The risk factors for refractory anaphylaxis are unknown. In the present analysis, we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.

  **Expert commentary:** According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and relatively safe treatment but may be insufficient. Therefore defined standardized treatment protocols for such cases of refractory anaphylaxis are needed to optimize the treatment. Point-of-care diagnostics may enable doctors to identify patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment. 
keywords: |
  anaphylaxis; epinephrine; glucagon; hypotension; methylene blue; shock
highlights: |
  These are the highlights. 
---

@@

# Introduction
Anaphylaxis is a potentially life-threatening mast cell-dependent reaction to most frequently food, drugs, insect venom. Anaphylaxis is a clinical diagnosis. It is highly likely if there is an acute onset of skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006].By contrast, the term refractory anaphylaxis has not established. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of minimum 300 µg adrenalin. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

@@

Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mastocyte activation syndromes) [@Worm2013]. Severe anaphylaxis is more likely to be elicited in older patients, patients with mastocytosis, and in temporal proximity to vigorous exercise (e.g. jogging) [@Worm2013]. Risk factors for _refractory_ anaphylaxis were not investigated previously.

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. To date, the acute treatment of anaphylaxis is uniquely recommended in international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are limited for ethical reasons. Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and papers not containing case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included 29 Papers reporting `r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

@@

An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked with increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].

@@

Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed (Table \@ref(tab:wholeTab)). We report the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.

Based on the analyzed reports, ambulatory patients were first treated with adrenaline, whereas patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, adrenaline followed by steroid and antihistaminic agents. Epinephrine was given in all cases in various doses.

@@

World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011]. Patients should be transported to a critical care unit. The AAAAI [@Lieberman] is suggesting a similar treatment but include dopamine, vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension. Also, the EAACI [@Muraro2014] guidelines suggest using glucagon in case of concomitant beta-blockade.

@@

The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce Nitric oxide (NO). NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000]. This in effect leads to the opening of the calcium ion channels in smooth muscle tissues and subsequent loss of vascular muscle contractility [@Omar2015].

Methylene blue is a well-investigated drug used for treating patients with methemoglobinemia [@Schirmer2003]. Buzato et al. investigated its properties as an inhibitor of nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue was mediated by the sGC inhibition. Therefore it is able to stop the vasoplegia mediated through the increased production of cGMP [@Buzato2005].

Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013]. Authors suggested that the inhibition of GC may result in the decrease of histamine and platelet-activating factor (PAF) production which is a potent mediator of anaphylaxis [@Moritoki1992].

The use of methylene blue in the current management of anaphylaxis is scarce. We received no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.

We propose to initiate the therapy in case of a refractory anaphylaxis with hypotension using 100 mg bolus i.v. and follow it with an infusion of 50 µg/kg/min. As a vector, a 100 mL of 5% dextrose can be used.

As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis. No side effects were reported due to methylene blue use in the above cases.

@@

In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be evaluated. Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol 
[`r which.sources(d$Source,"metara",list(d$What.helped,d$Therapy))`], methoxamine 
[`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`r which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
Drugs given perioperatively were the eliciting factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug. One patient reacted to systemic immune therapy (SIT) injection that was inadvertently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].

Use of vasopressin in refractory anaphylaxis was only suggested in management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

Seedat et al. described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia. The refractory shock was not responding to repeated adrenalin (total 2 mg), antihistaminics and hydrocortisone [@Seedat2014]. Rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when the biphasic reaction occurred. Authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014]. We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.

One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984]. Although this can be hardly available in critical care centers we would like to point out this minimally invasive method as a possible alternative measure in achieving normotension.

The European anaphylaxis register lists 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`). None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV. `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure. Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg=="no") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%

@@

Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lungs possess beta-adrenergic receptors.
Glucagon, a polypeptide hormone with potent inotropic and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomitant use of atenolol for arterial hypertension. They both reacted to contrast medium and after minutes they have developed severe anaphylaxis not responding to repeated doses of adrenaline. Infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].

Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014], therefore it is especially useful in patients who are chronically treated with beta-blockers. Patients with concomitant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade. Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon what, they discuss, would be a good therapeutic attempt in their case [@Laxenaire1984].

@@

A 15-year-old boy who was exposed to cow's milk underwent near-fatal anaphylaxis with refractory hypotension. After 150 minutes of resuscitative measures, a gastric drainage performed as a last resort treatment surprisingly led to a quick alleviation of anaphylaxis symptoms and patient recovered fully. Authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].

@@

In the most extreme cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort extracorporeal life support (ECLS) options. There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`]. Two patients reacted to chlorhexidine, one to rocuronium and in one case the elicitor could not be established. Only one of these patients was taking a beta blocker [@Yee2016]. All these patients received multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, corticosteroids, and antihistaminics. Upon prolonged resuscitation procedures two of these patients were placed on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device. After stabilizing of their condition, patients were weaned out from these supportive measures without neurological sequelae.

@@

We analyzed case reports of refractory anaphylaxis and evaluated the undertaken therapeutic approaches. The cases were most frequently elicited by drugs and happened during a medical procedure. All patients were treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v. and fluid replacement therapy.

>>>>

---
title: 'Risk Factors and Treatment of Refractory Anaphylaxis - a systematic review'
author:
  - Wojciech Francuzik
  - Sabine Dölle
  - Margitta Worm, MD
date: "`r format(Sys.time(), '%d.%m.%Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/template.docx"
bibliography: references.bib
csl: erci.csl
abstract: |
  **Introduction:** Patients experiencing anaphylaxis who do not recover after the treatment with intramuscular adrenaline are regarded to suffer from refractory anaphylaxis. The incidence of refractory anaphylaxis is estimated to range between 3--5% of anaphylaxis cases. The risk factors for refractory anaphylaxis are unknown.

  **Areas covered:** In the present analysis, we aimed to evaluate the management and risk factors of refractory anaphylaxis to highlight possible clinical implications for updating current management algorithms.

  **Expert commentary:** According to the international guidelines adrenaline given intramuscular (i.m.) is a rapid and safe treatment but may be insufficient. Therefore defined standardized treatment protocols for such cases of refractory anaphylaxis are needed to optimize the treatment. Point-of-care diagnostics may enable doctors to identify patients experiencing severe, refractory anaphylaxis early in order to initiate intensified critical care treatment. 
keywords: |
  anaphylaxis; epinephrine; glucagon; hypotension; methylene blue; shock
highlights: |
  These are the highlights. 
---

@@

# Introduction
Anaphylaxis is a potentially life-threatening, primarily mast cell-dependent reaction to most frequently food, drugs, insect venom. Anaphylaxis is a clinical diagnosis. It is highly likely if there is an acute onset of mainly skin symptoms in conjunction with cardiovascular symptoms like hypotension and/or respiratory compromise [@Sampson2006]. The pathogenesis of anaphylaxis is related to a sudden, systemic, (mostly IgE dependent) degranulation of mast cells or basophils leading to the release of multiple mediators like histamine, tryptase, chymase, platelet-activating factor, prostaglandins, and leukotrienes [@Houser2015].

Although anaphylaxis remains to be a clinical diagnosis [@Kemp2014], a standardized set of criteria to diagnose anaphylaxis have been established [@Sampson2006]. By contrast, the term refractory anaphylaxis has not been established yet. For the purpose of this review, we define refractory anaphylaxis as anaphylaxis (meeting the criteria by NIAID/FAAN  [@Sampson2006]) unresponsive to the treatment with at least 2 doses of a minimum 300 µg adrenaline. Unresponsiveness, in this case, is defined as a lack of expected normalization of clinical symptoms (i.e. rapid increase in blood pressure, bronchodilatation, tachycardia).

@@

Risk factors for anaphylaxis include age, food elicitors (cow's milk and hen's egg) in children and drug-induced anaphylaxis in adults as well as concomitant mastocytosis (or mast cell activation syndromes) [@Worm2013]. Severe anaphylaxis is more likely to be elicited in elderly patients, patients suffering from mastocytosis, and in temporal proximity to vigorous exercise (e.g. jogging) [@Worm2013]. Risk factors for _refractory_ anaphylaxis were not investigated previously.

The management algorithms for anaphylaxis are based on consensus expert recommendations and might be incomplete when it comes to reactions that are not responding to the first line therapy. To date, the acute treatment of anaphylaxis is uniquely recommended in the international guidelines and the mainstay of therapy is the early administration of intramuscular (i.m.) adrenaline [@Lazar2017].
However, an optimal anaphylaxis treatment in a given patient is difficult to study because it is an emergency condition and placebo-controlled studies are limited for ethical reasons. Therefore most of the knowledge regarding the treatment of anaphylaxis is based on expert recommendations which are graded as low-level evidence sources.

@@

We performed a database search in the "PubMed" database using the "RISmed"[@rismed] package for R statistical software [@R]. The search terms contained the words "refractory anaphylaxis" in articles published from 1950 till 2017. The search was conducted on 16.11.2017 and returned 150 results. All case reports of patients who experienced an anaphylactic event and did not respond to at least 600 µg of adrenaline were included in this study. Articles in English, Portuguese, French, and Spanish were included in the analysis, one article in Japanese was excluded. Additionally, case reports referenced in the included sources were added. We excluded reviews and articles not containing a case description as well as cases which documented an appropriate response to injected adrenaline. Also, cases describing mast cell activation syndrome and refractory vasoplegia without the definitive diagnosis of anaphylaxis were not included. We finally included `r unique(d$Sources) %>% length()` Papers reporting `r d$Author %>% length()` cases of anaphylaxis. The detailed inclusion flowchart is illustrated in Fig \@ref(fig:prismaflow).

@@

An early diagnosis of anaphylaxis during operative procedures can be suggested based on end-tidal CO~2~ collapse, which often is preceding cardiac arrest, even in the absence of any cutaneous manifestations. [@Baumann2009] In such case - treatment with adrenalin should be immediately initiated as delayed adrenaline use has been linked to increased mortality [@Simons2010] and refractoriness to adrenaline [@Fineman2015].

@@

Most of the treatment strategies were in accordance with the guidelines of anaphylaxis management [@Simons2011] but were supplemented by additional procedures that have saved patients' lives when first-line therapy failed (Table \@ref(tab:wholeTab)). We report on the most frequent procedures in order to evaluate them in the treatment of refractory anaphylaxis.

Based on the analyzed reports, out-patients were first treated with adrenaline, whereas patients experiencing anaphylaxis under anesthesia were usually first given intravenous volume replacement and 100% oxygen, adrenaline followed by steroid and antihistaminic agents. Epinephrine was given in all cases in various doses.

@@

The World Allergy Organisation guidelines on anaphylaxis management suggest that refractory anaphylaxis should be treated with intubation, ventilation, intravenous vasopressors, glucagon and anticholinergic drugs [@Simons2011]. Patients should be transported to a critical care unit. The AAAAI [@Lieberman] is suggesting a similar treatment but include dopamine, vasopressin, glucagon atropine, and methylene blue if epinephrine injections and volume expansion fail to alleviate hypotension. Also, the EAACI [@Muraro2014] guidelines suggest using glucagon in case of concomitant beta-blockade.

@@

The hypotension in anaphylaxis is mediated mainly through histamine [@Enjeti1983] which stimulates the endothelial nitric oxide synthase (eNOS) to produce nitric oxide (NO). NO subsequently activates the soluble guanylate cyclase (sGC) which is responsible for the production of cGMP out of guanosine triphosphate (GTP) [@Lucas2000]. This leads to the an opening of the calcium ion channels in smooth muscle tissues and the subsequent loss of vascular muscle contractility [@Omar2015].

Methylene blue is a well-investigated drug to treat patients with methemoglobinemia [@Schirmer2003]. Buzato et al. investigated its properties as an inhibitor of the nitric oxide-cyclic guanosine monophosphate (NO-cGMP) pathway in rabbits and suggested that the primary action of methylene blue is mediated by the sGC inhibition. Therefore it is able to stop the vasoplegia mediated through the increased production of cGMP [@Buzato2005].

Although the mechanism of action is thought to counteract the vasoplegia related hypotension, one case report successfully used methylene blue in the treatment of anaphylaxis without hypotension [@Bauer2013]. The authors suggested that the inhibition of GC may result in a decrease of histamine and platelet-activating factor (PAF) production which are potent mediators of anaphylaxis [@Moritoki1992].

Data about the use of methylene blue in the current management of anaphylaxis are scarce. We received no reports of methylene blue use in our International Registry of Anaphylaxis which contains over 10,000 cases to date.

In case of a refractory anaphylaxis with severe persisting hypotension we propose to apply a 100 mg bolus i.v. followed by an infusion of 50 µg/kg/min. As a vector, a 100 mL of 5% dextrose can be used.

As methylene blue does not block nitric oxide production its side effects are minimal (nausea, vomiting, abdominal pain, fever, hemolysis, hypotension, methemoglobinemia, arrhythmias, bluish skin, urine discol- oration, and hyperbilirubinemia [@Evora2007]). It may be considered a safe option for the treatment of anaphylaxis upon intensive care conditions. No side effects were reported due to methylene blue use in the reported-cases.

@@

In case of a hypotension refractory to adrenaline or volume replacement therapy -- vasopressors can be considered. Vasopressors were used in `r (d$t_vasopression==T) %>% sum` (`r (((d$t_vasopression==T) %>% sum)/((d$t_vasopression==T) %>% length)*100) %>% round(1)`%) of refractory cases. 
Vasopressin or its synthetic analogue terlipressin were used in `r sum(d$t_vasopressin)` cases. In 8 of these reports (`r (8/sum(d$t_vasopressin)*100) %>% round(1)`%) the injection of vasopressin reversed the refractory hypotension within minutes. 
Other vasopressors included metaraminol 
[`r which.sources(d$Source,"metara",list(d$What.helped,d$Therapy))`], methoxamine 
[`r which.sources(d$Source,"methox",list(d$What.helped,d$Therapy))`], dopamine 
[`r which.sources(d$Source,"dopam",list(d$What.helped,d$Therapy))`], and noradrenaline [`r which.sources(d$Source,"nor",list(d$What.helped,d$Therapy))`] infusions.
Drugs given perioperatively were the eliciting factors in all but two of these cases (`r round((length(d[temp.v,1])-2)/length(d[temp.v,1])*100,1)`%) and all patients reacted within minutes (range 5 - 20) upon exposition to the suspected drug. One patient reacted to systemic immune therapy (SIT) injection that was inadvertently extremely high-concentrated [@Bickell1984] and in one case the elicitor could not be established [@Heytman2004].

The use of vasopressin in refractory anaphylaxis was only suggested in the management recommendations from the AAAAI [@Lieberman] and was never vastly investigated. Dopamine, however, is recommended by 2 out of three guidelines. Nevertheless, dopamine might be related to higher a risk of mortality and adverse events than epinephrine or norepinephrine [@Avni2015]. Both norepinephrine and vasopressin may be appropriate second-line agents for the management of refractory hypotension. Vasopressin acts on the V1 receptors, while norepinephrine acts primarily on the alfa-adrenergic receptors, their synergistic effects may result in effective vasoconstriction [@DiChiara2008].

Seedat et al. described a case of anaphylaxis while undergoing a bilateral knee arthroscopy under general anesthesia. The refractory shock was not responding to repeated adrenaline (total 2 mg), antihistamines and hydrocortisone [@Seedat2014]. A rapid response was achieved with the last resort injection of 1 U ornipressin that was repeatedly effective when the biphasic reaction occurred. The authors concluded that vasopressin or ornipressin should be considered for the treatment of adrenaline-resistant anaphylactic shock before resuscitation is discontinued [@Seedat2014]. We support the use of vasopressin in refractory hypotension, combined with norepinephrine infusions.

One case report described the use of military anti-shock trousers to perform external vasopressin on a patient who received inadvertently high doses of systemic immune therapy allergen [@Bickell1984]. Although this can be hardly available in critical care centers we want to point out this minimally invasive method as a possible alternative measure in achieving normotension.

The European anaphylaxis register includes 18 cases where dopamine was used (mean age `r mean(dopam.anareg$d_age) %>% round(1)`; range `r dopam.anareg$d_age %>%  range() %>% paste0(collapse="--")`). None of these reactions were fatal and `r ((dopam.anareg$d_severity_rm %>% summary())[4] / dopam.anareg$d_severity_rm %>% summary() %>% sum()*100) %>% round(1)`% were Ring and Messmer grade IV. `r ((dopam.anareg$d_elicitor_gr5=="drugs") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`% of there reactions were elicited by drugs or occured during a medical procedure. Adrenaline was not used in `r ((dopam.anareg$d_522_adren_agg=="no") %>% sum()/length(dopam.anareg$d_elicitor_gr5)*100) %>% round(1)`%

@@

Adrenaline may fail to inhibit an anaphylactic event in patients taking beta-blockers as both heart and lung express beta-adrenergic receptors.
Glucagon, a polypeptide hormone with potent ino- and chronotropic actions was given in two cases of refractory hypotension in analyzed cases[@Zaloga1986; @Javeed1996]. Both patients receiving glucagon were undergoing refractory anaphylaxis with the concomittant use of atenolol for arterial hypertension. Both patients reacted to contrast medium and after minutes developed severe anaphylaxis not responding to repeated doses of adrenaline. An infusion of 1 mg glucagon per hour or 1 mg glucagon as a bolus relieved the refractory shock in both cases and it also successfully treated a recurrent biphasic reaction [@Javeed1996].

Adrenergic effects of glucagon are minimally influenced by beta-blockers [@Zaloga1986] as glucagon activates adenylyl cyclase directly [@Kemp2014] Therefore, it is especially useful in patients who are on long term beta-blocker treatments. Patients with concomittant beta-blockers often show bradycardia or normocardia during anaphylaxis [@Nag2017], as the reflexory tachycardia is suppressed by the beta-blockade. Authors who reported a fatal anaphylaxis case potentiated with beta blockers did not treat with glucagon but discuss this therapeutic option positively as well [@Laxenaire1984].

@@

A 15-year-old boy who was exposed to cow's milk underwent near-fatal anaphylaxis with refractory hypotension. After 150 minutes of resuscitative measures, a gastric drainage performed as a last resort treatment surprisingly led to a quick alleviation of anaphylactic symptoms and the patient recovered completely. These authors concluded that external gastric drainage should be considered an integral part of the treatment of severe life-threatening food-induced anaphylaxis [@Lazar2017].

@@

In the most severe cases, where all pharmacologic attempts fail to alleviate symptoms of anaphylaxis clinicians turned to last resort of extracorporeal life support (ECLS) options. There were 4 reports where patients reacted extremely severe and all were operated on because of a heart condition in general anesthesia [`r which.sources(d$Source,"ECLS",d$What.helped_gr)`]. Two patients reacted to chlorhexidine, one to rocuronium and in one case the elicitor could not be established. Only one of these patients was taking a beta blocker [@Yee2016]. All these patients received multiple doses of adrenaline (total dose ranged 1 -- 15 mg), vasopressors, high volume replacement therapy, corticosteroids, and antihistaminics. Upon prolonged resuscitation procedures two of these patients were placed on cardiopulmonary bypass (CPB) and two on an extracorporeal membrane oxygenation (ECMO) device. After stabilizing of their condition, patients were weaned off from these supportive measures without neurological sequelae.

@@

We analyzed case reports from literature and own data from our anaphylaxis registry of refractory anaphylaxis and critically evaluated the undertaken therapeutic approaches. These cases of refractory anaphylaxis were most frequently elicited by drugs and happened during a medical procedure. All patients were treated with adrenaline as a first line therapy, nearly all of them were given oxygen, steroids i.v. and fluid replacement therapy.
